---
title: rCharts
subtitle: Interactive JS Charts from R
author: Ramnath Vaidyanathan
github: {user: ramnathv, repo: rChartsNYT, branch: "gh-pages"}
framework: minimal
mode: selfcontained
widgets: [polycharts]
highlighter: highlight.js
hitheme: solarized_light
background: images/subtle_freckles.png
assets:
  css: 
    - "http://fonts.googleapis.com/css?family=Open+Sans"
    - "http://fonts.googleapis.com/css?family=Open+Sans+Condensed:700"
---

```{r echo = F, message = F, cache = F}
opts_chunk$set(results = 'asis', comment = NA, message = F, tidy = F)
require(rCharts)
options(RCHART_WIDTH = 600, RCHART_HEIGHT = 400)
read_chunk('app/ui.R'); read_chunk('app/server.R'); read_chunk('app/global.R')
```

# Replicating NY Times Interactive Graphic

This tutorial explains in detail, how I used `rCharts` to replicate this NY times [interactive graphic](http://www.nytimes.com/interactive/2013/03/29/sports/baseball/Strikeouts-Are-Still-Soaring.html?ref=baseball) on strikeouts in baseball. The end result can be seen [here](http://glimmer.rstudio.com/ramnathv/strikeouts) as a `shiny` application.

### Data

The first step is to get data on strikeouts by team across years. The NY Times graphic uses data scraped from [baseball-reference](http://www.baseball-reference.com/), using the `XML` package in R. However, I will be using data from the R package [Lahman](http://cran.r-project.org/web/packages/Lahman/index.html), which provides tables from [Sean Lahman's Baseball Database](http://www.seanlahman.com/baseball-archive/statistics/) as a set of data frames.

The data processing step involves using the [plyr](http://cran.r-project.org/web/packages/plyr/index.html) package to create two data frames:

1. `team_data` containing `SOG` (strikeouts per game) by `yearID` and team `name`
2. `league_data` containing `SOG` by `yearID` averaged across the league.


```{r}
require(Lahman)  
require(plyr)
dat = Teams[,c('yearID', 'name', 'G', 'SO')]
team_data = na.omit(transform(dat, SOG = round(SO/G, 2)))
league_data = ddply(team_data, .(yearID), summarize, SOG = mean(SOG))
```

### Charts

We will start by first creating a scatterplot of `SOG` by `yearID` across all teams. We use the `rPlot` function which uses the PolyChartsJS library to create interactive visualizations. The formula interface specifies the x and y variables, the data to use and the type of plot. We  also specify a `size` and `color` argument to style the points. Finally, we pass a `tooltip` argument, which is a javascript function that overrides the default tooltip to display the information we require. You will see below the R code and the resulting chart.


```{r}
require(rCharts)
p1 <- rPlot(SOG ~ yearID, data = team_data, type = 'point', 
  size = list(const = 2), color = list(const = '#888'), 
  tooltip="function(item){return item.SOG +'\n' + item.name + '\n' + item.yearID}"
)
p1$print('chart1')
```


Now, we need to add a line plot of the  average `SOG` for the league by `yearID`. We do this by adding a second layer to the chart, which copies the elements of the previous layer and overrides the `data`, `type`, `color` and `tooltip` arguments. The R code is shown below and you will note that the resulting chart now shows a blue line chart corresponding to the league average `SOG`.


```{r}
p1$layer(data = league_data, type = 'line', 
  color = list(const = 'blue'), copy_layer = T, tooltip = NULL)
p1$print('chart2')
```

Finally, we will overlay a line plot of `SOG` by `yearID` for a specific team `name`. Later, while building the shiny app, we will turn this into an input variable that a user can choose from a dropdown menu. We use the layer approach used earlier and this time override the `data` and `color` arguments so that the line plot for the team stands out from the league average.


```{r}
myteam = "Boston Red Sox"
p1$layer(data = team_data[team_data$name == myteam,], color = list(const = 'red'),
  copy_layer = T)
p1$print('chart3')
```

Let us add a little more interactivity to the chart. To keep it simple, we will use handlers in PolychartJS to initiate an action when a user clicks on a point. Here is where the magic of `knitr` shines (thanks to [yihui/knitr](yihui/knitr)), as we can mix coffeescript code in our document. The current handler is a simple one, which just displays the name of the team clicked on. If you are familiar with Coffeescript, the code is self explanatory.

```{r interactivity, engine='coffee', echo = T, eval = F}
graph_chart3.addHandler (type, e) ->
  data = e.evtData
  if type == 'click'
    alert "You clicked on the team: " + data.name.in[0]
```

<script>
```{r ref.label = 'interactivity', engine = 'coffee', echo = F, eval = T}
```
</script>


### Application

Now it is time to convert this into a Shiny App. We will throw the data processing code into `global.R` so that it can be accessed both by `ui.R` and `server.R`. For the dropdown menu allowing users to choose a specific team, we will restrict the choices to only those which have data for more than 30 years. Accordingly, we have the following `global.R`.

```{r eval = F}
## global.R
require(Lahman)  
require(plyr)
dat = Teams[,c('yearID', 'name', 'G', 'SO')]
team_data = na.omit(transform(dat, SOG = round(SO/G, 2)))
league_data = ddply(team_data, .(yearID), summarize, SOG = mean(SOG))
THRESHOLD = 30
team_appearances = count(team_data, .(name))
teams_in_menu = subset(team_appearances, freq > THRESHOLD)$name
```

For the UI, we will use a bootstrap page with controls being displayed in the sidebar. Shiny makes it really easy to create a page like this. See the annotated graphic below and the `ui.R` code that accompanies it to understand how the different pieces fit together.

<a href='http://glimmer.rstudio.com/ramnathv/strikeouts'>
  <img src="http://www.clipular.com/c?5151018=o2zGtsIzD20s1dp25X1mLRSUTMk&f=e04b448074961bf4efcb426b91886d8b">
</a>

```{r ref.label = 'ui.R', eval = F}
```

We now need to write the server part of the shiny app. Thankfully, this is the easiest part, since it just involves wrapping the charting code inside `renderChart` and replacing user inputs to enable reactivity. We add a few more lines of code to set the height and title and remove the axis titles, since they are self explanatory.


```{r ref.label = 'server.R', eval = F}
```

### Notes

This post is fully reproducible. You will need to have the `dev` version of the [slidify](http://github.com/ramnathv/slidify) package installed.


